A full approximation storage multilevel method \\ for nonlinear variational inequalities

### 2023-05-21 paper/exact/plap1d.tex
include solution figures, and improve text

For the purposes of building a nonlinear variational inequality verification case we will construct exact solutions to the one-dimensional case of the \emph{$p$-Laplacian obstacle problem}.  Over an open set $\Omega\subset \RR^d$, for $p>1$, $f \in L^1(\Omega)$, and $\psi \in W^{1,p}(\Omega)$, recall that the problem is to find $u(x)$ satisfying
\begin{equation}
-\Div\left(|\grad u(x)|^{p-2} \grad u(x)\right) = f(x) \qquad \text{and} \qquad u(x) \ge \psi(x). \label{plap}
\end{equation}

The weak form of this problem is well-posed with Dirichlet boundary conditions, and in particular there is a unique solution $u \in \mathcal{X} = W_0^{1,p}(\Omega)$ \cite{KinderlehrerStampacchia1980}.  Equivalently, the coercive functional $J[u] = \int_\Omega \frac{1}{p} |\grad u|^p - f u$ has a unique minimum over $\mathcal{K}=\{v \ge \psi\} \subset \mathcal{X}$, a closed and convex subset.  Observe that if $p=2$ then the operator is the ordinary Laplacian, if $1<p<2$ then $D=|\grad u|^{p-2}>0$ is unbounded as $|\grad u| \to 0$ (\emph{fast diffusion}), while if $p>2$ then $D\to 0$ as $|\grad u| \to 0$ (\emph{degenerate diffusion}).  The degenerate case violates uniform ellipticity, despite coercivity.

We will solve a one-dimensional case of \eqref{plap}, with
\begin{equation}
f(x) = \begin{cases} \alpha, & |x| < 1 \\ -\alpha, & |x| > 1\end{cases} \qquad \text{and} \qquad \psi(x) = - \beta |x| \label{data}
\end{equation}
on the interval $\Omega = (-3,3) \subset \RR^1$, for $\alpha > 0$ and $\beta \ge 0$.
% FIXME  I think the cases where \beta < 0$ are fine too, though eventually the free boundary will be outside $(-3,3)$
That is, we seek $u(x) \ge -\beta |x|$ so that
\begin{equation}
-\left(|u'(x)|^{p-2} u'(x)\right)' = \alpha \mathbbm{1}_{(-1,1)}(x) - 2 \mathbbm{1}_{(-3,-1)\cup(1,3)}(x).  \label{concrete}
\end{equation}

@@

We will set parameters $\alpha=1$ and $\beta = 0.2$ for convenience, and numerically test powers $p=3/2,2,4$, corresponding to $q=3,2,4/3$, respectively.  Note that for our chosen parameters $\alpha,\beta$, the requirement $\beta < \alpha^{q-1}/q$ is satisfied in the range $p \in [3/2,4]$.

>>>>

For the purposes of building a nonlinear variational inequality verification case we will construct exact solutions to the one-dimensional case of the \emph{$p$-Laplacian obstacle problem}.  Over an open set $\Omega\subset \RR^d$, for $p>1$, $f \in L^1(\Omega)$, and $\psi \in W^{1,p}(\Omega)$, recall that the problem is to find $u(x) \ge \psi(x)$ satisfying
\begin{equation}
-\Div\left(|\grad u(x)|^{p-2} \grad u(x)\right) = f(x) \qquad \text{on the set where } u(x) > \psi(x). \label{plap}
\end{equation}

The weak form of this problem is well-posed with Dirichlet boundary conditions, and in particular there is a unique solution $u \in \mathcal{K}=\{v \ge \psi\} \subset W_0^{1,p}(\Omega)$ \cite{KinderlehrerStampacchia1980}.  Equivalently, the coercive functional $J[u] = \int_\Omega \frac{1}{p} |\grad u|^p - f u$ has a unique minimum over $\mathcal{K}$, a closed and convex subset.  Observe that if $p=2$ then the operator is the ordinary Laplacian.  If $1<p<2$ then $D=|\grad u|^{p-2}>0$ is unbounded as $|\grad u| \to 0$ (\emph{fast diffusion}), while if $p>2$ then $D\to 0$ as $|\grad u| \to 0$ (\emph{degenerate diffusion}).  The degenerate case violates uniform ellipticity, despite coercivity.

We will solve a one-dimensional case of \eqref{plap}, with
\begin{equation}
f(x) = \begin{cases} \alpha, & |x| < 1 \\ -\alpha, & |x| > 1\end{cases} \qquad \text{and} \qquad \psi(x) = - \beta |x| \label{data}
\end{equation}
on the interval $\Omega = (-3,3) \subset \RR^1$, for $\alpha > 0$ and $\beta \ge 0$.
% FIXME  I think the cases where \beta < 0$ are fine too, though eventually the free boundary will be outside $(-3,3)$
That is, we seek $u(x) \ge -\beta |x|$ so that
\begin{equation}
-\left(|u'(x)|^{p-2} u'(x)\right)' = \alpha \left(\mathbbm{1}_{(-1,1)}(x) - \mathbbm{1}_{(-3,-1)\cup(1,3)}(x)\right).  \label{concrete}
\end{equation}

@@

The $\beta=0$ case, in which the obstacle $\psi(x)=0$ is flat, implies that the solution has zero gradient at the free boundary ($u'(\xi)=0$).  Note that this implies a loss of uniform ellipticity at the free boundary in the $p>2$ degenerate diffusion cases, which makes the obstacle problem very challenging numerically, so we avoid that difficulty by setting $\beta>0$.

For numerical tests we set parameters $\alpha=1$ and $\beta = 0.2$, and consider powers $p=3/2,2,4$, corresponding to $q=3,2,4/3$ respectively.  (For these $\alpha,\beta$ values, the requirement $\beta < \alpha^{q-1}/q$ is satisfied when $p > 5/4$.)  Numerical and exact solutions for the $p=3/2$ and $p=4$ cases are shown in Figure \ref{fig:solutions}.

\begin{figure}
\mbox{\includegraphics[width=0.45\textwidth]{solution1p5.pdf} \qquad \includegraphics[width=0.45\textwidth]{solution4.pdf}}
\caption{Low-resolution $p$-Laplacian obstacle problem solutions for $\alpha=1$, $\beta=0.2$, and $p=1.5$ (left) or $p=4$ (right).}
\label{fig:solutions}
\end{figure}
